Floating-Point Arithmetic on Deterministic Systems

### 2024-01-12 mss.tex
Post.
Post a partial revision based on review.

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[hyphens]{url}

\usepackage{fontspec}
\setmainfont{Linux Libertine O}
\newfontfamily\monofont{Urbit Sans Mono TT SemiBold}
\renewcommand{\setmonofont}{\monofont}
\renewcommand{\texttt}[1]{{\monofont #1}}

\usepackage{listings}
\lstset{
    basicstyle=\setmonofont\small,
    showspaces=false
}
\usepackage{titlesec}
\newfontfamily\urbitfont{Urbit Sans TT SemiBold}
\titleformat{\section}{\urbitfont\Large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\urbitfont\Large\bfseries}{\thesubsection}{1em}{}
\usepackage{titling}
\renewcommand{\maketitlehooka}{\urbitfont\bfseries}

%% Configure first-page footer and later-page header.
\usepackage{fancyhdr}
\setlength{\headheight}{15.2pt}
\pagestyle{fancy}
%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \texttt{\textasciitilde lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{Floating-Point Arithmetic on Deterministic Systems}
\author{Neal E. Davis \texttt{\textasciitilde lagrev-nocfep}}

@@

Modern digital computers deal, at their root, in binary representation, entirely zeros and ones.\footnote{Analog computers may operate on a continuum of value.  Computer logic systems may be architected on other numeric bases for their logic, such as the 1837 Analytical Engine's decimal system and the 1958 Сетунь Setun's ternary system.}  These are often formally considered to be whole numbers in a number base of two.  However, numerical calculations very frequently require the use of numbers with a fractional part to adequately represent the elements of a computation.

Early numeric computing tended to focus on problems of interest to military and national security applications, such as thermonuclear reaction cascades and neutron transport calculations.  Such calculations typically involve arrays of many steps, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed even greater prominence with the rise of three-dimensional gaming, with a particular emphasis on speed over correctness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  Deterministic computing in particular requires that a given computation be reproducible exactly.  This includes, for Urbit as a state machine, that the event log replay across platforms to the same result.

@@

Binary computer values are at root easily represented as nonnegative integers.  However, it is frequently convenient when working with human applications to either use other numeric bases (notably decimal and hexadecimal) or to permit non-integer mathematics.

In the historical development of mathematics, logical problems in each set of numbers drove the \{discovery, invention\} of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition $+$ or multiplication $\times$ produces a value within the set; however, permitting subtraction $-$ of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division $/$ similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis.  https://plato.stanford.edu/entries/pythagoras/} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the real numbers $\mathbb{R}$.  The real numbers $\mathbb{R}$ are continuous, meaning (among other things) that any pair of values has a difference, however small.

Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes have been proposed to mitigate this situation by permitting a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as ENIAC and MANIAC were employed in the 1940s for solving arrays of thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  \citep[p.~3]{Kahan1997a}} and floating-point representations.

The basic concept of floating point arithmetic is that it permits the representation of a discrete subset of $\mathbb{R}$ by composing a significand, a base, and an exponent.  The significand is the set of significant digits, possibly including the sign; the base is the understood number base (typically 2); and the exponent is the power to which that base is put before multiplying by the significand to yield the result.\footnote{Compare so-called “engineering” notation, such as \texttt{1e5} for 10,000, which compactly represents the significand 1 and the exponent 5 with an understood base of 10 indicated by \texttt{e} and a sign, implicit for positive.}  The most ubiquitous floating-point format in use today is defined by the IEEE 754 standard, but certain hardware platforms such as GPGPUs utilize alternative floating-point arithmetic representations.\footnote{We cite bfloat16 \citep{Wang2019} and TensorFloat-32 \citep{Kharva2020}, among others.}

To summarize, given an abstract description of a floating point system, there are several practical implementations that can be derived.  We need to specify at least four quantities:

\begin{enumerate}
  \item  Sign ($+$/$-$)
  \item  Significand (magnitude)
  \item  Base
  \item  Exponent
\end{enumerate}

\noindent
The base is presumably fixed by the protocol, leaving three free values for the implementation to encode.

@@

IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but not leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with IEEE 754 is exact even if all arguments are.”  The ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citep[fn.~21]{ISO9899}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

Since the IEEE 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as \texttt{<<} left shift and \texttt{+} addition do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

IEEE 754 floating-point arithmetic (and its predecessors) has some significant mathematical compromises even in its formal specification.\footnote{None of this intends to demean the impressive technical accomplishment of IEEE 754 and its architects.  “\emph{Nani gigantum humeris insidentes}” (Bernard of Chartres).}  For instance, as a result of the discrete nature of the bitwise representation in \texttt{E} and \texttt{F}, floating-point mathematics are actually a subset of discrete mathematics masquerading as real mathematics.  This has nontrivial consequences for certain aspects of calculations, including error accrual.  In particular, three facts dominate the resolution:

@@

Truncation error results from terminating repeating “binaries” (by analogy with “decimals”).  Just as $\frac{1}{3} = 0.\bar{3} \approx 0.3333…$ has a finite precision when written in base-10, numbers that are not precise powers of two result in repeating fractions.  These necessarily terminate at the resolution of the significand.  The number and nature of truncation and rounding can significantly affect the accuracy of floating-point arithmetic and algorithms \citep{Izquierdo2006}.

@@

Formally neither associative nor commutative for the above reasons, floating-point arithmetic can break our mathematical intuitions in interesting ways.  However, this is a consistent and well-understood phenomenon.  For our purposes as designers of deterministic computers, the most damning indictment has to do not with IEEE 754 itself but with manufacturer deviation in hardware implementation.  In 1997, William Kahan himself complained (\emph{justamente}) about the compromises inherent in the standard for compiler implementers:

@@

\begin{enumerate}
  \item  Optional, discretionary, or advisory aspects
  \item  Gaps or omissions in the specification
  \item  Failure to implement the specification exactly
  \item  Out-of-sequence computations
\end{enumerate}

@@

Failure to implement IEEE 754 would most likely happen inadvertently (as with the Pentium FDIV bug in the 1990s).  Chipset designers may also deviate from this behavior.

@@

As a further example, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citep[p.~5]{Kahan1997}.\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification raises the specter of nondeterminism.}

@@

\lstset{language=Hoon}
\begin{lstlisting}
++  add
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)
  ?:  =((cmp:si w x) --1)
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}

@@

Why can't we just allow different results in the last binary places of the significand?  Philosophically, Urbit holds the following statements as bedrock truth:\footnote{\citet{Monk2020}}

@@

Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat \citep{Foote1999}.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity:\footnote{\citet{Monk2020a}}

@@

\noindent
In this hypothetical case,“any extended precision bits will be lost in the first calculation of a+b when it is assigned to x. The result of the second calculation of a+b may be held in a working register at the extended precision and potentially contain additional value bits not held in x, the result of the equality test then being false.”

@@

K \& R C permitted the compiler to re-order floating-point expressions by associativity, which could run afoul of our limitations.  ANSI C (C89), recognizing the issue introduced by this innocuous change, forbade such re-ordering \citep{MacDonald1991}.  Compiler optimizations (e.g. GCC's \texttt{-O3}) can bypass this restriction, once again breaking determinism;\footnote{This can be mitigated again in turn by the use of the \texttt{volatile} designation, but this is sufficient to illustrate the problem.} for instance, floating-point operations can be pipelined, leading to out-of-order execution.

The fly in the ointment for Urbit is that deterministic computing requires deterministic computing; that is, jet-accelerated Nock equivalents must reliably produce the same results (both to each other and to Nock) regardless of the runtime on which it is being evaluated.  Thus even small irregularities in floating-point implementations have macroscopic ramifications for deterministic computing.  Any guarantee broken breaks them all, just as it would for a formal correctness proof.\footnote{I was once asked by a retired computer science professor if such guarantees would make things easier.  Well, at the end developer level!}

The challenge of the lack of determinacy for certain critical applications has been acknowledged before, such as by James Demmel and the \texttt{ReproBLAS} team \citep{Demmel2017, ReproBLAS} and by \citeauthor{Dawson2013}.  Dawson makes much of the effect of rounding modes and the option to disable subnormals, both of which would have major effects on computational reproducibility.  The situation is in fact worse for transcendental functions, because there is necessarily truncation and/or rounding error

(The situation is in fact worse for transcendental functions \citep{Dawson2013}.)
% Conversion between bases – such as _printf(“%1.8e”);_ – is not guaranteed to be identical across all implementations. Doing perfect conversions efficiently was an unsolved problem when the original IEEE standard came out and while it has since been solved this doesn’t mean that everybody does correctly rounded printing

@@

\begin{enumerate}
  \item  Hardware-supported floating-point arithmetic
  \item  Software-defined floating-point library
  \item  Opaque calculations
  \item  Stored results
  \item  Proscribing IEEE 754
\end{enumerate}

@@

As outlined above, execution of software-equivalent floating-point computations produced from source by different compilers on different hardware architectures may lead to small differences in outcome, nonnegligible for a deterministic computer.  Thus, for this and a constellation of related reasons, hardware-supported floating-point arithmetic seems to be \emph{prima facie} unviable for deterministic computing.\footnote{Although not of grave consequence, the C language (as of C23) does not implement at least two types specified by IEEE 754-2019 and recent predecessors:  \texttt{binary128} quadruple precision and \texttt{binary256} octuple precision.  While neither are significant losses, we also note that Urbit does not currently support a C-style \texttt{long double} type.  C's \texttt{long double} is 80 bits wide on some common consumer hardware, such as the x86-64 architecture, but is 128 bits wide on the 64-bit ARM architecture.  (The situation is worse for Python, whose \texttt{numpy.float128} type eponymously advertises itself as quadruple precision but is in fact a regular 80-bit \texttt{long double}.)  Some compilers and libraries do support quadruple-precision floating-point mathematics, such as GCC's \texttt{\_\_float128} type.  We note that IEEE 754 80-bit extended-precision could be implemented using the \texttt{++fn} core should demand arise.}

@@

If you controlled the compiler and runtime execution stack to sufficient precision, could you yield deterministic floating-point arithmetic from the hardware?  “A translator that generates very high performance code is of no use if the final behavior is incorrect” \citep[p.~189]{Jones2008}.

To start off, what must be considered part of the stack in this sense?  At a minimum, the compiler and linker toolchain (including flags and options) and the actual runtime must be included.  (This introduces a dependence between Martian software on Earthling software, repugnant to the Urbit ethos.)

@@

Can the C-defined floating-point environment (as supplied by \texttt{fenv.h}) answer to this need?  This augurs the ability to specify not only rounding modes and access floating-point exception status flags, but it is not clear that this environmental control spans the entire output of floating-point computations.\footnote{“The floating-point environment access and modification is only meaningful when \texttt{\#pragma STDC FENV\_ACCESS} is set to \texttt{ON}.  \ldots{} In practice, few current compilers, such as HP aCC, Oracle Studio, and IBM XL, support the \texttt{\#pragma} explicitly, but most compilers allow meaningful access to the floating-point environment anyway.” \citep{CPPRef}}

@@

An optimized portable deterministic software library for floating-point calculations may be a sufficiently fast solution to meet Urbit's needs even for vector computations.  A different avenue worthy of investigation is to take IEEE 754 compliant floating-point values as inputs and outputs, then transform into a local representation for an optimized portable deterministic calculation.  For instance, \citet{Thall2007} presents the concept of “unevaluated sums”, a generalized technique for accruing error in situations where additional precision is necessary for accuracy.  However, even with an agreed-upon standard library like SoftFloat, it is important to keep in mind that results like transcendental functions are still not correctly known.\footnote{To correctly calculate a trigonometric function for \texttt{double} may take over a hundred bits of precision before correct rounding can be determined.  Furthermore, the C \texttt{math.h} implementation of $\sin$ may or may not use \texttt{fsin}.}  This particular poses a problem for functions like $\sin$ which may be calculated by different routes in Hoon/Nock and in C/Rust.  For the time being, we conclude that Urbit's discipline requires only using Hoon/Nock implementations of transcendental functions.

@@

To the second, we observe that although Urbit is a state machine whose history is part of its state, in practice we can mitigate event log growth by either \emph{choppping} the event log by storing its state and permitting replay forward from that point or \emph{tombstoning} data which should never be available again.\footnote{The memory implications of these are not necessary here, but take place in different arenas:  the runtime versus the Arvo noun arena.}

@@

Instead of repeating computations that have been made in the past, what if we cached the result of all of them, so that any new computations with the same values are guaranteed to result in the same value via a cache lookup instead of a calculation?

\begin{quote}
A recently proposed hardware acceleration technique is to store the results of previous multiplication and division operations in a cache, reusing rather than recalculating the result whenever possible. (Dynamic profiling has found that a high percentage of these operations share the same operands as previous operations.)  \citep[p.~1148]{Jones2008}\footnote{Cf. Jones' citation of \citeauthor{Citron1998}.}
\end{quote}

@@

\begin{enumerate}
  \item  \texttt{\%real}, an IEEE 754 floating-point value
  \item  \texttt{\%uint}, an unsigned integer
  \item  \texttt{\%int2}, a twos-complement signed integer
  \item  \texttt{\%cplx}, a BLAS-compatible packed \texttt{float}
  \item  \texttt{\%unum}, a unum/posit value
  \item  \texttt{\%fixp}, a fixed-precision value
\end{enumerate}

\noindent
for which \texttt{\%real} allows the rounding mode to be specified; and \texttt{\%fixp} requires an additional metadata statement of the expected precision.

>>>>

\usepackage{ustj}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{Floating-Point Arithmetic on Deterministic Systems}
\author{N. E. Davis~\patp{lagrev-nocfep} \\ Urbit Foundation}
\date{}

@@

Modern digital computers deal, at their root, in binary representation, entirely zeros and ones.\footnote{Analog computers may operate on a continuum of value.  Computer logic systems may be architected on other numeric bases for their logic, such as the 1837 Analytical Engine's decimal system and the 1958 Setun's (Сетунь) ternary system.}  These are often formally considered to be whole numbers in a number base of two.  However, numerical calculations very frequently require the use of numbers with a fractional part to adequately represent the elements of a computation.

Early numeric computing tended to focus on problems of interest to military and national security applications, such as partial differential equations and numerical optimization.  Such calculations typically involve arrays, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed even greater prominence with the rise of three-dimensional gaming, with a particular emphasis on speed over correctness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  Error accrual in such cases is deterministic, but

On the other hand, deterministic computing describes the ability for a given computation be reproducible exactly.  Such reproducibility permits referential transparency and more powerful reasoning about a program's results and dependencies.  This includes, for Urbit as a state machine, that the event log replay across platforms to the same result.

@@

Binary computer values are at root easily represented as non-negative integers.  However, it is frequently convenient when working with human applications to either use other numeric bases (notably decimal and hexadecimal) or to permit non-integer mathematics.

In the historical development of mathematics, logical problems in each set of numbers drove the discovery and elaboration of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition ($+$) or multiplication ($\times$) produces a value within the set; however, permitting subtraction ($-$) of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division ($/$) similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis.  https://plato.stanford.edu/entries/pythagoras/} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the set of real numbers.  The set of real numbers, denoted by $\mathbb{R}$, is characterized by its continuity, implying that for any two distinct values within this set, there exists a difference, no matter how small.

Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as ENIAC and MANIAC were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  \citep[p.~3]{Kahan1997a}} and floating-point representations.

The basic concept of floating point arithmetic is that it permits the representation of a discrete subset of $\mathbb{R}$ by composing a significand, a base, and an exponent.  The significand is the set of significant digits, possibly including the sign; the base is the understood number base (typically 2); and the exponent is the power to which that base is put before multiplying by the significand to yield the result.\footnote{Compare so-called “engineering” notation, such as \texttt{1e5} for 10,000, which compactly represents the significand 1 and the exponent 5 with an understood base of 10 indicated by \texttt{e} and a sign, implicit for positive.}  The most ubiquitous floating-point format in use today is defined by the IEEE 754 standard, but certain hardware platforms such as GPUs utilize alternative floating-point arithmetic representations.\footnote{We cite \texttt{bfloat16} \citep{Wang2019} and TensorFloat-32 \citep{Kharva2020}, among others.}

To summarize, given an abstract description of a floating point system, there are several practical implementations that can be derived.  We need to specify at least four quantities:  sign,\footnote{We could omit the sign by introducing an offset or only allowing positive values.} significand, base, and exponent.\footnote{The exponent has a bias so that it in turn does not need a sign.}  The base is presumably fixed by the protocol, leaving three free values for the implementation to economically encode.

@@

IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with IEEE 754 is exact even if all arguments are.”  The ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citep[fn.~21]{ISO9899}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

Since the IEEE 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as left shift (\texttt{<<}) and addition (\texttt{+}) do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

IEEE 754 floating-point arithmetic and its predecessors have some significant mathematical compromises even in its formal specification.\footnote{None of this intends to demean the impressive technical accomplishment of IEEE 754 and its architects.}  For instance, as a result of the discrete nature of the bitwise representation in \texttt{E} and \texttt{F}, floating-point mathematics are actually a subset of discrete mathematics masquerading as real mathematics.  This has non-trivial consequences for certain aspects of calculations, including error accrual.  In particular, three facts dominate the resolution:

@@

Truncation error results from terminating repeating “binaries” (by analogy with “decimals”).  Just as $\frac{1}{3} = 0.\bar{3} = 0.3333…$ has a finite precision when written in base-10, numbers that are not precise powers of two result in repeating fractions.  These necessarily terminate at the resolution of the significand.  The number and nature of truncation and rounding can significantly affect the accuracy of floating-point arithmetic and algorithms \citep{Izquierdo2006}.

@@

Formally neither associative nor commutative for the above reasons, floating-point arithmetic can break our mathematical intuitions in interesting ways.  However, this is a consistent and well-understood phenomenon.  For our purposes as designers of deterministic computers, the most damning indictment has to do not with IEEE 754 itself but with manufacturer deviation in hardware implementation.  In 1997, William Kahan himself complained (justly) about the compromises inherent in the standard for compiler implementers:

@@

\begin{enumerate}
  \item  Optional, discretionary, or advisory aspects.
  \item  Gaps or omissions in the specification.
  \item  Failure to implement the specification exactly.
  \item  Out-of-sequence computations.
\end{enumerate}

@@

Failure to implement IEEE 754 correctly may happen inadvertently, as with the Pentium FDIV bug in the 1990s \citep{Edelman1997}.  Alternatively, chipset designers may deviate from the specification for reasons of performance or limitations in the architecture.

@@

As a further example, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citep[p.~5]{Kahan1997}.
%\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification raises the spectre of nondeterminism.}

@@

%\lstset{language=Hoon}
\begin{lstlisting}
++  add
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)
  ?:  =((cmp:si w x) --1)
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}

@@

Why can't we just allow different results in the last binary places of the significand?  Philosophically, Urbit holds the following statements as bedrock truth \citep{Monk2020}:

@@

Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat \citep{Foote1999}.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity \citet{Monk2020a}:

@@

\noindent
In this hypothetical case, “any extended precision bits will be lost in the first calculation of a+b when it is assigned to x. The result of the second calculation of a+b may be held in a working register at the extended precision and potentially contain additional value bits not held in x, the result of the equality test then being false.”

@@

K\&R C permitted the compiler to re-order floating-point expressions by associativity, which could run afoul of our limitations.  ANSI C (C89), recognizing the issue introduced by this innocuous change, forbade such re-ordering \citep{MacDonald1991}.  Compiler optimizations (e.g. GCC's \texttt{-O3}) can bypass this restriction, once again breaking determinism;\footnote{This can be mitigated in turn by the use of the \texttt{volatile} designation, but this is sufficient to illustrate the problem.} for instance, floating-point operations can be pipelined, leading to out-of-order execution.

The fly in the ointment for Urbit's deterministic computing is that jet-accelerated Nock equivalents must reliably produce the same results (both to each other and to Nock) regardless of the runtime on which it is being evaluated.  Thus even small irregularities in floating-point implementations have macroscopic ramifications for deterministic computing.  Any guarantee broken breaks them all, just as it would for a formal correctness proof.\footnote{I was once asked by a retired computer science professor if such guarantees would make things easier.  Well, at the end developer level!}

The challenge of the lack of determinacy for certain critical applications has been acknowledged before, such as by James Demmel and the \texttt{ReproBLAS} team \citep{Demmel2017, ReproBLAS} and by \citeauthor{Dawson2013}.  Dawson makes much of the effect of rounding modes and the option to disable subnormals, both of which would have major effects on computational reproducibility.  The situation is worse for transcendental functions, because there is necessarily truncation and/or rounding error \citep{Dawson2013}.
% Conversion between bases – such as _printf(“%1.8e”);_ – is not guaranteed to be identical across all implementations. Doing perfect conversions efficiently was an unsolved problem when the original IEEE standard came out and while it has since been solved this doesn’t mean that everybody does correctly rounded printing

@@

\begin{enumerate}
  \item  Hardware-supported floating-point arithmetic.
  \item  Software-defined floating-point library.
  \item  Opaque calculations.
  \item  Stored results.
  \item  Proscribing IEEE 754.
\end{enumerate}

@@

As outlined above, execution of software-equivalent floating-point computations produced from source by different compilers on different hardware architectures may lead to small differences in outcome, non-negligible for a deterministic computer.  Thus, for this and a constellation of related reasons, hardware-supported floating-point arithmetic seems to be \emph{prima facie} unviable for deterministic computing.\footnote{Although not of grave consequence, the C language (as of C23) does not implement at least two types specified by IEEE 754-2019 and recent predecessors:  \texttt{binary128} quadruple precision and \texttt{binary256} octuple precision.  While neither are significant losses, we also note that Urbit does not currently support a C-style \texttt{long double} type.  C's \texttt{long double} is 80 bits wide on some common consumer hardware, such as the x86-64 architecture, but is 128 bits wide on the 64-bit ARM architecture.  (The situation is worse for Python, whose \texttt{numpy.float128} type eponymously advertises itself as quadruple precision but is in fact a regular 80-bit \texttt{long double}.)  Some compilers and libraries do support quadruple-precision floating-point mathematics, such as GCC's \texttt{\_\_float128} type.  We note that IEEE 754 80-bit extended-precision could be implemented using the \texttt{++fn} core should demand arise.}

@@

If you controlled the compiler and runtime execution stack to a sufficient degree, could you yield deterministic floating-point arithmetic from the hardware?  “A translator that generates very high performance code is of no use if the final behavior is incorrect” \citep[p.~189]{Jones2008}.

To start off, what must be considered part of the stack in this sense?  At a minimum, the compiler and linker toolchain (including flags and options) and the actual runtime must be included.  (This explicitly introduces a dependence between Martian software and Earthling software, repugnant to the Urbit ethos.)

@@

Can the C-defined floating-point environment (as supplied by \texttt{fenv.h}) answer to this need?  This affords the ability to specify not only rounding modes and access floating-point exception status flags, but it is not clear whether this environmental control portably spans the entire output of floating-point computations.\footnote{“The floating-point environment access and modification is only meaningful when \texttt{\#pragma STDC FENV\_ACCESS} is set to \texttt{ON}.  \ldots{} In practice, few current compilers, such as HP aCC, Oracle Studio, and IBM XL, support the \texttt{\#pragma} explicitly, but most compilers allow meaningful access to the floating-point environment anyway.” \citep{CPPRef}}

@@

An optimized portable deterministic software library for floating-point calculations may be a sufficiently fast solution to meet Urbit's needs even for vector computations.  A different avenue worthy of investigation is to take IEEE 754 compliant floating-point values as inputs and outputs, then transform into a local representation for an optimized portable deterministic calculation.  For instance, \citet{Thall2007} presents the concept of “unevaluated sums”, a generalized technique for accruing error in situations where additional precision is necessary for accuracy.  However, even with an agreed-upon standard library like SoftFloat, it is important to keep in mind that exact floating-point results for transcendental functions are still not correctly known in many cases.\footnote{To correctly calculate a trigonometric function for \texttt{double} may take over a hundred bits of precision before correct rounding can be determined.  Furthermore, the C \texttt{math.h} implementation of $\sin$ may or may not use \texttt{fsin}.}  This particular poses a problem for functions like $\sin$ which may be calculated by different routes in Hoon/Nock and in C/Rust.  For the time being, we conclude that Urbit's discipline requires only using Hoon/Nock implementations of transcendental functions.

@@

To the second, we observe that although Urbit is a state machine whose history is part of its state, in practice we can mitigate event log growth by either \emph{chopping} the event log by storing its state and permitting replay forward from that point or \emph{tombstoning} data which should never be available again.\footnote{The memory implications of these are not necessary here, but take place in different arenas:  the runtime versus the Arvo noun arena.}

@@

Instead of repeating computations that have been made in the past, what if we cached the result of all of them, so that any new computations with the same values are guaranteed to result in the same value via a cache lookup instead of a calculation?  Urbit uses memoization frequently in Arvo and in the runtime, so this is an aesthetically compatible option; we consider its feasibility.

\begin{quote}
A recently proposed hardware acceleration technique is to store the results of previous multiplication and division operations in a cache, reusing rather than recalculating the result whenever possible. (Dynamic profiling has found that a high percentage of these operations share the same operands as previous operations.)  \citep[p.~1148]{Jones2008}\footnote{Cf. Jones' citation of \citeauthor{Citron1998}} %no period for et al.
\end{quote}

@@

\begin{enumerate}
  \item  \texttt{\%real}, an IEEE 754 floating-point value
  \item  \texttt{\%uint}, an unsigned integer
  \item  \texttt{\%int2}, a twos-complement signed integer
  \item  \texttt{\%cplx}, a BLAS-compatible ordered pair
  \item  \texttt{\%unum}, a unum/posit value
  \item  \texttt{\%fixp}, a fixed-precision value
\end{enumerate}

\noindent
for which \texttt{\%real} allows the rounding mode to be specified; \texttt{\%cplx} consists of a pair of two values, real and imaginary parts; and \texttt{\%fixp} requires an additional metadata statement of the expected precision.
